%% publication list
\noindent
\pagestyle{empty}
%\selectlanguage{nenglish}

\section*{Declaration}

The dissertation or parts of it were \textbf{previously published}. I attach information on previous publications according to \S 2 (4) of the \emph{Promotionsordnung} of TU Berlin and regulations of Faculty II.

\vspace{2ex}

This is a list of the peer-reviewed publications containing parts of the dissertation and the detailed contributions of the co-authors to each publication:

\begin{enumerate}

    \item \textsc{C. Minelli, R. Garcia-Diez, A. Sikora, C. Gollwitzer, M. Krumrey, A. Shard}, Characterization of IgG-protein-coated polymeric nanoparticles using complementary particle sizing techniques, \emph{Surface and Interface Analysis} \textbf{46} 663-667 (2014)

        \hspace{10pt} AS and MK conceived the study in the framework of the BioSurf project. CM and AS collected and analysed the DCS and DLS data. RGD and CG performed the SAXS measurements and RGD evaluated the experimental data. CM drafted the manuscript. All authors read, approved and contributed to the final manuscript.

    \item \textsc{R. Garcia-Diez, C. Gollwitzer, M. Krumrey}, Nanoparticle characterization by continuous contrast variation in small-angle {{X}}-ray scattering with a solvent density gradient, \emph{Journal of Applied Crystallography} \textbf{48}, 20-28 (2015)

        \hspace{10pt} RGD, CG and MK designed the study and developed the methodology. RGD and CG performed the measurements. RGD analysed the data and drafted the manuscript. All authors read, approved and contributed to the final manuscript.

    \item \textsc{R. Garcia-Diez, C. Gollwitzer, M. Krumrey, Z. Varga}, Size determination of a liposomal drug by small-angle X-ray scattering using continuous contrast variation, \emph{Langmuir} \textbf{32 (3)}, 772-778 (2016)

        \hspace{10pt} RGD and ZV designed the study and developed the methodology. RGD, ZV and CG performed the measurements. RGD analysed the data. RGD and ZV drafted the manuscript. All authors read, approved and contributed to the final manuscript.
        
    \item \textsc{R. Garcia-Diez, A. Sikora, C. Gollwitzer, C. Minelli, M. Krumrey}, Simultaneous size and density determination of polymeric colloids by continuous contrast variation in small angle {{X}}-ray scattering, \emph{European Polymer Journal} \textbf{81} 641–649 (2016) 

        \hspace{10pt} RGD and CM designed the study and developed the methodology. RGD and CG performed the SAXS measurements. CM and AS collected and evaluated the DCS data. RGD analysed the SAXS data and drafted the manuscript. All authors read, approved and contributed to the final manuscript.

\end{enumerate}

\vspace{3cm}

\noindent Berlin, den 15. Februar 2017 \hfill Raül Garc\'{i}a Diez

\cleardoublepage
